---
title: "nyc flight data exploration ,time series and regression model"
author: "Aritra Banerjee"
date: "2024-03-28"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

``
**installing important libraries**
```{r warning=FALSE,massage=FALSE}
library(reshape2)
library(tidyverse)
library(ggplot2)
library(GGally)
library(caret)
library(dplyr)
library(forecast)
```
**DATA LOADING AND EXPLORATION**
```{r warning=FALSE,massage=FALSE,results='hide',include=FALSE}
flights<-read.csv("nycflights.csv")
#exploring the dataset
head(flights)
summary(flights)
names(flights)
glimpse(flights)
```
**DATA CLEANING**
```{r warning=FALSE,massage=FALSE,results='hide',include=FALSE}
#Checking for na values
sum(is.na(flights))
#removing na values
flights<-na.omit(flights)

```
 **EXPLORATORY DATA ANALYSIS**
```{r echo=FALSE,warning=FALSE,massage=FALSE,results='hide'}
#creating data frame
data <- data.frame(
  DepartureDelay = flights$dep_delay,
  ArrivalDelay = flights$arr_delay,
  FlightDistance = flights$distance
)


#HISTOGRAM PLOT
ggplot(data, aes(x = DepartureDelay)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Departure Delays", x = "Departure Delay (minutes)", y = "Frequency") 

ggplot(data, aes(x = ArrivalDelay)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Arrival Delays", x = "Arrival Delay (minutes)", y = "Frequency") 

ggplot(data, aes(x = FlightDistance)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Flight Distances", x = "Flight Distance (miles)", y = "Frequency") 

#UNDERSTANDING DATA USING BOXPLOT
ggplot(flights,aes(x=factor(month), y=dep_delay))+geom_boxplot()+labs(x="months",y="departure delay")

#SUMMARIZING MORE FACTORS

flights %>%
  select(dep_delay) %>%
  summary()

flights %>% 
  group_by(month) %>% 
  summarise(mean_dd=mean(dep_delay)) %>% 
  arrange(desc(mean_dd))


flights %>% 
  group_by(month) %>% 
  summarise(median_dd=median(dep_delay)) %>% 
  arrange(desc(median_dd))



#flight delay rate
flights<-flights %>% 
  mutate(arrival_type=ifelse(arr_delay<=0,"ON TIME","DELAYED"))
flights<-flights %>% 
  mutate(departure_type=ifelse(dep_delay<=0,"ON TIME","DELAYED"))

flights %>% select(departure_type,arrival_type) %>% table
#proportion of flight on time arrival
(3508/(9291+3508))*100

#flight delay visualization
flights<-flights %>% 
  mutate(dept_type=ifelse(dep_delay<5,"ON TIME","DELAYED"))

flights %>% 
  group_by(origin) %>% 
  summarise(on_time_dept_rate=sum(dept_type=="ON TIME")/n()) %>% 
  arrange(desc(on_time_dept_rate))

ggplot(data=flights,aes(x=origin,fill=dept_type))+geom_bar()

#SPEED ANALYSIS
flights<-flights %>% mutate(avg_speed=distance/(air_time/60))
speed_data<-flights %>% select(avg_speed,tailnum) %>% filter(avg_speed == max(avg_speed)) 
ggplot(flights,aes(x=distance,y=avg_speed))+geom_point()



```

CONCLUSION:- The distribution of departure delays of all flights is right skewed signifying the data to be unimodal. Also, right skewness is proved by the summary obtained in which mean > median of the departure delay distribution.
July in which there is highest average departure delay of flights from NYC airport.
LGA  is one of the three major NYC airports that has a better on time percentage for departing flights.
Plane with tail number N666DN having the highest average speed.
27.40% of the flights arrived on time though departed late.

**CORRELATIONAL ANALYSIS**
```{r  }
corrlation_matrix<-cor(select(flights,dep_delay,arr_delay,distance))
corrlation_matrix
## Visualize the correlation matrix as a heatmap
ggplot(melt(corrlation_matrix), aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "yellow", midpoint = 0,
                       limit = c(-1,1), space = "Lab", name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1)) +
  coord_fixed()

#FINDING CORRELATION VIUALIZATION
ggpairs(data)
```
CONCLUSION:- From the correlation matrix we can conclude that Arrival delay and departure delay has a high positive corelation
distance and departure delay has low negative correlation and so for arrival delay.

**CATEGORICAL VARIBALBE ANALYSIS**
```{r }
flights_dataframe<- data.frame(
  Airline = flights$carrier,
  Destination = flights$dest,
  Departure_Delay =flights$dep_delay)

average_delays <- flights_dataframe %>%
  group_by(Airline, Destination) %>%
  summarise(Avg_Delay = mean(Departure_Delay))

average_delays<- melt(average_delays, id.vars = c("Airline", "Destination"))
#VISUALI
ggplot(average_delays, aes(x = Destination, y = Airline, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Average Departure Delay Across Airlines and Destinations",
       x = "Destination",
       y = "Airline",
       fill = "Average Departure Delay") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed(ratio=4,expand=T) 


```
conclusion:-From the plot we can clearly see that the airline UA has max average arrival delay.
**TIME SERIES ANALYSIS**
```{r }
#TIME SERIES ANALYSIS
#by month
monthly_delays <- flights %>%
  group_by(month) %>%
  summarize(avg_delay = mean(dep_delay))
#weekly
flights$day_of_week <- weekdays(as.Date(paste(flights$year, flights$month, flights$day, sep = "-")))
weekly_delays <- flights %>%
  group_by(day_of_week) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE))

#BY DAY
daily_delays <- flights %>%
  group_by(day) %>%
  summarize(avg_delay = mean(dep_delay))
#weekly delay trend
ggplot(weekly_delays, aes(x = reorder(day_of_week, avg_delay), y = avg_delay)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Departure Delay by Day of the Week",
       x = "Day of the Week",
       y = "Average Departure Delay (minutes)")
# Monthly Delay Trend
ggplot(monthly_delays, aes(x = month, y = avg_delay)) +
  geom_line(colour="red") +
  labs(title = "Average Departure Delay by Month", x = "Month", y = "Average Delay (minutes)")

# Daily Delay Trend
ggplot(daily_delays, aes(x = day, y = avg_delay)) +
  geom_line(color="blue") +
  labs(title = "Average Departure Delay by Day", x = "Day", y = "Average Delay (minutes)")


```
**AIRPORT ANALYSIS**
```{r  }
#flight delay 
flights<-flights %>% 
  mutate(dept_type=ifelse(dep_delay<5,"ON TIME","DELAYED"))

#Origin Wise Delay 
flights %>% 
  group_by(origin) %>% 
  summarise(on_time_dept_rate=sum(dept_type=="ON TIME")/n()) %>% 
  arrange(desc(on_time_dept_rate))
#visualizing the delay of diff airport
ggplot(data=flights,aes(x=origin,fill=dept_type))+geom_bar()
```
conclusion:-From the plot we can visualize that EWR has the maximum departure delay and LGA has the minimun departure delay
** STATISTICAL TEST**
```{r}
#two sample test
JFK_delay<-flights %>% filter(origin=="JFK") %>% select(arr_delay)
JFK_delay
LGA_delay<-flights %>% filter(origin=="LGA") %>% select(arr_delay)
LGA_delay
t_test<-t.test(JFK_delay,LGA_delay)
t_test

```

